---
title: "Generating the harmonized NSCH dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generating the harmonized NSCH dataset}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette shows how to produce the harmonized NSCH dataset (survey years
2016 through 2024) with the `nsch` package: installing the package, getting the
raw data, generating the combined dataset, pulling a subset of variables, and
adding a raw variable that the package does not harmonize.

The code chunks are not run when the vignette builds, because the pipeline reads
the raw NSCH Stata files, which are not shipped with the package. Run the same
calls locally against your own copy of the data.

## Installing the package

The package lives on GitHub. Install it with `remotes` (or `pak`):

```{r}
# install.packages("remotes")
remotes::install_github("NAU-ASD3/nsch")
```

If the repository is private, you will need a GitHub personal access token with
repo read access, either set as the `GITHUB_PAT` environment variable or passed
directly:

```{r}
remotes::install_github("NAU-ASD3/nsch", auth_token = "your_pat_here")
```

```{r}
library(nsch)
library(data.table)
```

## Getting the data and generating the dataset in one step

The pipeline reads the original NSCH topical Stata files. The simplest path is to
let the package download them for you and generate the harmonized dataset in a
single call, by passing `download = TRUE`:

```{r}
d <- get_clean_data(years = 2016:2024, download = TRUE)
```

This fetches each year's file from the NSCH data portal into
`NSCH_data/00_original_Stata` (relative to your working directory), then runs the
full harmonization pipeline. The download happens once; on later runs the files
are already present, so you can drop `download = TRUE`.

If you already have the Stata files, point `data.path` at them and skip the
download:

```{r}
d <- get_clean_data(
  years = 2016:2024,
  data.path = "path/to/your/NSCH_data/00_original_Stata")
```

Either way the result is a single long `data.table` with a `year` column, the
harmonized variables from the package's configuration, and the survey unique
identifier `hhid`. Categorical variables come back as labeled factors, with the
value labels stored as the factor levels, so you can read and tabulate them
directly.

```{r}
dim(d)
d[, .N, by = year]
```

## Working with a subset of variables

The full pull returns every variable in the package's desired-variable list. If
you only need a few, select them afterward. Emergency-room visits (`hospitaler`)
is a good example: it is one of the harmonized variables, so it is already in the
output as a labeled factor (`None`, `1 time`, `2 or more times`).

```{r}
er <- d[, .(year, hhid, hospitaler)]
er[, .N, by = .(year, hospitaler)][order(year, hospitaler)]
```

A note on cross-year comparability: `hospitaler` does not use the same response
categories every year. Through 2021 it was asked as None / 1 / 2 or more; from
2022 on it splits the top into 2-3 and 4 or more. The package emits the native
per-year categories; reconciling them across years is left to your analysis. The
year-level audit in the analysis repository lists every variable with this kind of
shift.

## Adding a variable the package does not harmonize

The package harmonizes a curated set of variables. The raw Stata files contain
many more. You can pull one of those extra variables into your dataset, but it
comes through as-is: it is not renamed, merged, transformed, given factor labels,
or checked for cross-year consistency the way the curated variables are. You are
responsible for confirming it behaves sensibly across years.

BMI classification (`bmiclass`) is an example. It is in the raw files but not in
the harmonized set. Read each raw year, keep the survey ID (`hhid`) and
`bmiclass`, stack the years, and join onto the harmonized data.

```{r}
"bmiclass" %in% names(d)   # FALSE: not harmonized

add_bmiclass <- function(years, data.path = "NSCH_data/00_original_Stata") {
  files <- get_all_years(data.path = data.path, years = years)
  parts <- lapply(seq_len(nrow(files)), function(i) {
    raw <- read_nsch_dta(files[["dta.path"]][i])
    raw[, .(hhid, year = files[["year"]][i], bmiclass)]
  })
  data.table::rbindlist(parts)
}

bmi <- add_bmiclass(2016:2024)
d_plus <- merge(d, bmi, by = c("hhid", "year"), all.x = TRUE)
```

The join is on `hhid` and `year`, which together uniquely identify a row, so the
row count is unchanged.

Because `bmiclass` skipped harmonization, it comes through as its raw numeric
codes (not a labeled factor like the harmonized variables), and you should check
it yourself before using it. Its native codes are 1 = less than 5th percentile,
2 = 5th to less than 85th, 3 = 85th to less than 95th, 4 = at or above 95th, with
values in the 990s marking missing data. Two things worth confirming for any added
variable: that the codes mean the same thing in every year, and that the
missing-data sentinels are handled the way you expect.

```{r}
d_plus[, .N, by = .(year, bmiclass)][order(year, bmiclass)]
```

In this case `bmiclass` happens to use the same codes in every year, so it is
straightforward to use. That is not guaranteed for an un-harmonized variable; it
is exactly the kind of consistency the pipeline handles for you on the curated
set, and that you have to verify yourself when you step outside it.

## Changing the variable set through the configuration

The approach above adds a variable at the analysis stage, which is usually the
simplest. If you would rather have the pipeline itself carry a different set, the
desired-variable list lives in the package configuration
(`inst/extdata/variable-config.json`, the `desired_variables` array). Adding a
name that already has harmonization rules includes it in every `get_clean_data()`
run; removing a name drops it. A raw variable added to that list without matching
rules still comes through un-harmonized, the same caveat as above.

For a one-off subset, filtering the returned `data.table` (as in the `hospitaler`
example) is simpler and does not require touching the package.
